---
title: "Ontario NOS"
author: "Danielle Ethier"
date: "11/03/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

Updated: January 2026

Code used to create the basic data summaries Birds Canada's Nocturnal Owl Survey in Ontario. Used for annual reporting to MNRF. 

##Load Packages and Create Directories
**Search and change year to current analysis year**

```{r LoadPackages, include=FALSE}

#install.packages("remotes")
#remotes::install_github("BirdStudiesCanada/naturecounts")

library (naturecounts)
library (tidyverse)
library (reshape)

#mapping
library(sf)
library(rnaturalearth)
library(ggmap)
library(mapview)

#create and use data and output directories
dir.create("Data/2026") 
dir.create("Output/2026") 

out.dir <- paste("Output/2026") 
dat.dir <- paste("Data/2026") 

```

##Data Downloas

ONOWLS collection is by request if you are accessing it through the naturecounts R package. 

You need to sign up for a naturecounts account https://www.birdscanada.org/birdmon/default/register.jsp, and make a request for the data here https://www.birdscanada.org/birdmon/default/searchquery.jsp. 

Once you have received notice that the data are ready, you will use your username, password, and request_id to download the data into R. 

**Important** if you would like to access the environmental covariates, then change the 'fields_set' to 'extended'. Otherwise you can access the count data with the default setting. 

```{r download data, echo=FALSE}

#sample script will not work without your username, password, and request_id

ONOWLS_dat <- nc_data_dl(collections = "ONOWLS",
                      request_id = 000000, username = "USER",
                      info = "MY REASON", fields_set="extended")

#save to local drive for safe keeping
write.csv(ONOWLS_dat, "Data/2026/ONOWLS_TODAYSDATE.csv")

#read data from file
ONOWLS_dat<-read.csv("Data/2026/ONOWLS_TODAYSDATE.csv")

```

##Data Manipulation

If you downloaded the extend BMDE you now have 285 columns of data. Many of which will not be useful. Here we select the fields that most researchers will be interested in using. 

EffortUnit and ObservationDescriptor describe the data for each Measurement and Count, respectively. 

```{r select data, echo=FALSE}

in.data<-ONOWLS_dat %>% filter(RouteIdentifier!="`")

#Get Unique location name
in.data$Location<-sub("-?stop ?[[:alnum:]]+$", "", in.data$Locality,
                        ignore.case = TRUE)

in.data<-in.data %>% dplyr::select(SamplingEventIdentifier, SurveyAreaIdentifier,RouteIdentifier, Location, SiteCode, collection, survey_day, survey_month, survey_year, survey_week, protocol_id, CollectorNumber, Collector,  EffortUnits1, EffortMeasurement1, EffortUnits2, EffortMeasurement2, EffortUnits3, EffortMeasurement3, EffortUnits4, EffortMeasurement4, EffortUnits5, EffortMeasurement5, EffortUnits6, EffortMeasurement6, EffortUnits7, EffortMeasurement7, EffortUnits8, EffortMeasurement8, EffortUnits9, EffortMeasurement9, EffortUnits10, EffortMeasurement10, EffortUnits11, EffortMeasurement11, EffortUnits12, EffortMeasurement12, EffortUnits14, EffortMeasurement14, species_id, CommonName, ScientificName, latitude, longitude, bcr, StateProvince, ObservationDescriptor, ObservationCount, ObservationDescriptor2, ObservationCount2,ObservationDescriptor3, ObservationCount3, ObservationDescriptor4, ObservationCount4, ObservationDescriptor5, ObservationCount5)

```

##Data Summaries

A few quick summaries are created here for data checking and end-of-year reporting. 

```{r summarize data, echo=FALSE}

#summarize the number of route run in each year
route.yr.tot<-in.data %>% group_by(survey_year) %>% summarise(nroute=n_distinct(RouteIdentifier))

#summarize the number of routes run in each year by Protocol
#protocol_id 22=Northern 36=Central
route.yr<-in.data %>% group_by(survey_year, protocol_id) %>% summarise(nroute=n_distinct(RouteIdentifier))

route.yr<-cast(route.yr, survey_year~ protocol_id, value=c("nroute"))

collect.yr<-in.data %>% group_by(survey_year, protocol_id) %>% summarise(ncollectors=n_distinct(CollectorNumber))

collect.yr<-cast(collect.yr, survey_year~protocol_id, value="ncollectors")

write.csv(route.yr, "Output/2026/ONOWLS.RoutesPerYear.csv")
write.csv(collect.yr, "Output/2026/ONOWLS.CollectorsPerYear.csv")

#summarize the total count of each species per year
in.data$ObservationCount<-as.numeric(in.data$ObservationCount) 
species.yr<-in.data %>% group_by(survey_year, CommonName, protocol_id) %>% summarise(count=sum(ObservationCount))
species.yr<-cast(species.yr, CommonName+protocol_id~survey_year, value="count")
write.csv(species.yr, "Output/2026/ONOWLS.SpeciesPerYear.csv")

```

Creating some summaries that can be used for volunteer appreciated. 

Number of years a survey routes was run (includes Collector)

```{r route run}
long_route<-in.data %>% filter(RouteIdentifier!="`")

long_route<-long_route %>% group_by(protocol_id, CollectorNumber, Collector, RouteIdentifier) %>% summarise(min_year = min(survey_year), max_year=max(survey_year), n_year=n_distinct(survey_year)) 

write.csv(long_route, "Output/2026/ONOWLS.CountRouteYears_collector.csv")
```

Number of years a survey routes was run (excludes Collector)

```{r route run}
long_route2<-in.data %>% filter(RouteIdentifier!="`")

long_route2<-long_route2 %>% group_by(protocol_id, RouteIdentifier) %>% summarise(min_year = min(survey_year), max_year=max(survey_year), n_year=n_distinct(survey_year)) 

write.csv(long_route2, "Output/2026/ONOWLS.CountRouteYears.csv")
```

Number of years a survey routes was run (don't include Collector + Location Name)

```{r route run}
long_route3<-in.data %>% filter(RouteIdentifier!="`")

long_route3<-long_route3 %>% group_by(protocol_id, RouteIdentifier, Location) %>% summarise(min_year = min(survey_year), max_year=max(survey_year), n_year=n_distinct(survey_year)) 

write.csv(long_route3, "Output/2026/ONOWLS.CountRouteYears_location.csv")
```

Number of years a volunteer participated. 

```{r volunteer}
vol_year<-in.data %>% filter(RouteIdentifier!="`")

vol_year<-vol_year %>% group_by(CollectorNumber, Collector)%>% summarise(min_year = min(survey_year), max_year=max(survey_year), n_year=n_distinct(survey_year)) 

write.csv(vol_year, "Output/2026/ONOWLS.VolunteerYears.csv")

```

##Mapping routes

##TO DO
##MAPPING IN LEAFLETS FOR INTERACTIVE DISPLAY
Output to use include so far the CountRouteYear.csv

##ADD TRENDS TO EACH ROUTE FOR EACH SPECIES

Visualize the routes for each survey year. This is done to ensure that routes are not falling outside Ontario. 

```{r map data, echo=FALSE}

register_stadiamaps("1cdeb586-3f5a-4ef9-813b-02375efa9f21") 

plot.data<-in.data %>% select(RouteIdentifier, survey_year, latitude, longitude) %>% distinct() 

#Create a spatial dataframe

plot.data<-na.omit(plot.data)
plot_sf <- st_as_sf(plot.data, coords = c("longitude", "latitude"), crs = 4326)

#Create the base layer map
map <- get_stadiamap(bbox = as.numeric(st_bbox(plot_sf)), zoom = 5)

#Create a new plot for each year
plot<-ggmap(map) + 
  geom_point(data = plot.data, aes(x = longitude, y = latitude))+
  facet_wrap( ~ survey_year, ncol=5)

print(plot)


##Annual Map

library(rnaturalearth)
library(sf)
library(dplyr)
library(ggplot2)


# Canada provinces
can_prov <- ne_states(country = "Canada", returnclass = "sf")
on_sf <- can_prov %>%
  filter(name_en == "Ontario") %>%
  st_transform(4326)

pts_2025 <- in.data %>%
  filter(survey_year == 2025) %>%
  distinct(RouteIdentifier, SiteCode, latitude, longitude, protocol_id)

pts_2025<-pts_2025 %>% mutate("SurveyArea" = ifelse(protocol_id == 22, "North", ifelse(protocol_id == 36, "Central", "Ontario")))

p<-ggplot() +
  geom_sf(data = on_sf, fill = "grey95", colour = "grey70", linewidth = 0.3) +
  geom_point(
    data = pts_2025,
    aes(x = longitude, y = latitude, colour = factor(SurveyArea)),
    size = 2, alpha = 0.8
  ) +
  coord_sf() +  # no xlim/ylim: full Ontario polygon
  scale_color_brewer(palette = "Dark2", name = "Protocol") +
  labs(x = "Longitude", y = "Latitude", title = "Nocturnal Owl Survey Routes 2025") +
  theme_classic()

pdf(paste(plot.dir, "Route2025_Owls.pdf", sep=""))
plot(p)
while(!is.null(dev.list())) dev.off()  

```


**When using the data for scientific research it is important that zero-filling is take into consideration. There is a helper function built into the naturecounts package: format_zero_fill. Addition documentation can be found here:https://birdstudiescanada.github.io/naturecounts/articles/format-zero-fill.html **
