---
title: "02_DataManip"
author: "Danielle Ethier"
date: "01/02/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updated: Feb 2025

Code used to manipulate the data for old trend analysis. This is in need of updating. Terribly inefficient!

**Search and replace year to current analysis year**

##Load Packages and Create Directories

```{r LoadPackages, include=FALSE}

#install.packages("remotes")
#remotes::install_github("BirdStudiesCanada/naturecounts")

library (naturecounts)
library (tidyverse)
library (reshape)
library(lubridate)

#mapping
library(sf)
library(rnaturalearth)
library(ggmap)
library(mapview)

out.dir <- paste("Output/2024")
dat.dir <- paste("Data/2024")

```

Download the saved data from the last session  

```{r downloaddata, echo = FALSE, message = FALSE}

in.data<-read.csv("Data/2024/ONOWLS_13Feb.csv")

in.data<-in.data %>% filter(RouteIdentifier!="`") %>% drop_na(RouteIdentifier)

```

Prepare files for analysis

```{r prepareowlsfiles, echo = FALSE, message = FALSE}

#owls data select appropriate columns from database

#note that the old code removed duplicated and also used the survey window filed. This data is not transferred into the BMDE table. Duplicates are not removed. There is a work around for survey window.  

owls<-in.data %>% dplyr::select(protocol_id, species_id, CommonName, SamplingEventIdentifier, RouteIdentifier, statprov_code, SurveyAreaIdentifier, YearCollected, MonthCollected, DayCollected, ObservationCount, DecimalLatitude, DecimalLongitude, Remarks2)

#create a station ID column
owls<-owls %>% separate(SamplingEventIdentifier, c("del1", "del2", "stn"), sep="-", remove=FALSE) %>% dplyr::select(-del1, -del2)

#drop NA in dates
owls<-owls %>% drop_na("YearCollected", "MonthCollected", "DayCollected", "SurveyAreaIdentifier")

#add day of year (doy) to dataframe to be used later 
owls <- owls %>%
  mutate(date = ymd(paste(YearCollected, MonthCollected, DayCollected, sep = "/")),
         doy = yday(date))

#There should only be one SurveryAreaIdentifier per year. If there are multiple then we need to select the last survey_day for the year. Do this using the slice max doy. 
owls<-owls %>% group_by(SurveyAreaIdentifier, YearCollected) %>% slice(which.max(doy)) %>% ungroup()

#need to remove duplicates which are in the Remarks field of the BMDE then the remarks field can be safely removed. In othewords, keep the NAs. First we need to create a new level for the factor and add the level 0. Then replace NA with 0 and filter for 0. 
owls$Remarks2[is.na(owls$Remarks2)] <- 0
owls<-owls %>% filter(Remarks2=="0")
```

Create a sampling event table to zero fill the dataframes 

```{r zerofilldata, echo = FALSE, message = FALSE}

stndata<- owls %>% dplyr::select(SamplingEventIdentifier, RouteIdentifier, stn, protocol_id, YearCollected, MonthCollected, DayCollected) %>% distinct ()

```

#Old owls analysis

```{r speciesleveldata, echo = FALSE, message = FALSE}

owlss<-owls %>% dplyr::select(SamplingEventIdentifier, YearCollected, MonthCollected, DayCollected, CommonName)

##This need re-written and put into a loop. 

nswo=subset(owlss, CommonName=="Northern Saw-whet Owl")
nswo$nowls<-1 #for counting the number of owls 
barr=subset(owlss, CommonName=="Barred Owl")
barr$nowls<-1
boow=subset(owlss, CommonName=="Boreal Owl")
boow$nowls<-1
ggow=subset(owlss, CommonName=="Great Gray Owl")
ggow$nowls<-1
ghow=subset(owlss, CommonName=="Great Horned Owl")
ghow$nowls<-1

#Merge each species with the stndata dataframe to zero fill

###NSWO###
nsworoute=merge(nswo, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(nsworoute)

nsworoute$CommonName[is.na(nsworoute$CommonName)]="Northern Saw-whet Owl" 
nsworoute <- subset(nsworoute, select = c(CommonName,YearCollected, RouteIdentifier,stn, nowls));head(nsworoute)####only the variables we want
nsworoute$nowls[is.na(nsworoute$nowls)] <- 0
summary(nsworoute)
raw_nswo<-nsworoute

nsworoute<-nsworoute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
nsworoute$spp<-"NSWO"

###BARR###
barrroute=merge(barr, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(barrroute)

barrroute$CommonName[is.na(barrroute$CommonName)]="Barred Owl" 
barrroute <- subset(barrroute, select = c(CommonName,YearCollected,RouteIdentifier,stn, nowls));head(barrroute)####only the variables we want
barrroute$nowls[is.na(barrroute$nowls)] <- 0
summary(barrroute)
raw_barr<-barrroute

barrroute<-barrroute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
barrroute$spp<-"BARR"

###BOOW###
boowroute=merge(boow, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(boowroute)

boowroute$CommonName[is.na(boowroute$CommonName)]="Boreal Owl" 
boowroute <- subset(boowroute, select = c(CommonName,YearCollected,RouteIdentifier,stn, nowls));head(boowroute)####only the variables we want
boowroute$nowls[is.na(boowroute$nowls)] <- 0
summary(boowroute)
raw_boow<-boowroute

boowroute<-boowroute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
boowroute$spp<-"BOOW"
summary(boowroute)

###GGOW###
ggowroute=merge(ggow, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(ggowroute)

ggowroute$CommonName[is.na(ggowroute$CommonName)]="Great Gray Owl" 
ggowroute <- subset(ggowroute, select = c(CommonName,YearCollected,RouteIdentifier,stn, nowls));head(ggowroute)####only the variables we want
ggowroute$nowls[is.na(ggowroute$nowls)] <- 0
summary(ggowroute)
raw_ggow<-ggowroute

ggowroute<-ggowroute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
ggowroute$spp<-"GGOW"
summary(ggowroute)

###GHOW###
ghowroute=merge(ghow, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(ghowroute)

ghowroute$CommonName[is.na(ghowroute$CommonName)]="Great Horned Owl" 
ghowroute <- subset(ghowroute, select = c(CommonName,YearCollected,RouteIdentifier,stn, nowls));head(ghowroute)####only the variables we want
ghowroute$nowls[is.na(ghowroute$nowls)] <- 0
summary(ghowroute)
raw_ghow<-ghowroute

ghowroute<-ghowroute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
ghowroute$spp<-"GHOW"
summary(ghowroute)

#combine all species again
owlroute<-rbind(nsworoute,ggowroute,ghowroute,barrroute,boowroute)
raw_owldata<-rbind(raw_nswo,raw_boow,raw_ggow,raw_barr,raw_ghow)
head(raw_owldata)
owlroute$yr=as.factor(owlroute$YearCollected) #changes yr in dataframe to factor
owlroute$route=as.factor(owlroute$RouteIdentifier)
```

```{r routeleveldata, echo = FALSE, message = FALSE}
#In the process change the dataframe name to owldata

#protocol_id 22=NORTHERN 36=CENTRAL
region <- in.data %>% dplyr::select(SamplingEventIdentifier, RouteIdentifier, protocol_id, DecimalLatitude, DecimalLongitude) %>% distinct() %>% drop_na()

#create a statation ID column
region<- region %>% separate(SamplingEventIdentifier, c("del1", "del2", "stn"), sep="-", remove=FALSE) %>% dplyr::select(-del1, -del2)

#remove leading zeros by changing route from character to integer
region$protocol_id[region$protocol_id=="22"]<-"NORTH"
region$protocol_id[region$protocol_id=="36"]<-"CENTRAL"
#change some column names
colnames(region)[which(names(region) == "protocol_id")] <- "region"
colnames(region)[which(names(region) == "DecimalLatitude")] <- "lat"
colnames(region)[which(names(region) == "DecimalLongitude")] <- "long"
#select the first stop on the route as the routes lat long
region<-region %>% filter (stn == "1") %>% dplyr::select(-stn, -SamplingEventIdentifier) %>% distinct()

#combine with the owlroute data
owldata<-left_join(owlroute, region, by="RouteIdentifier")

#identify the regions with missing lat long for Kathy to fix. 
na_dropcoord<-subset(owldata, is.na(owldata$long))
owldata<-owldata %>% drop_na(lat, long)
```

#remove COVID data
```{r filter years}

owldata<-owldata %>% filter(YearCollected!=2020)

```

Write raw datatable for MNRF Report

```{r writedata, echo = FALSE, message = FALSE}

write.csv(owldata, "Data/2024/owldata_2024.csv")
owldata<-read.csv("Data/2024/owldata_2024.csv")

```

Subset data by species and region for analysis

```{r writedata, echo = FALSE, message = FALSE}
#create a subset of NSWO in CENTRAL region
nswoc=subset(owldata, spp=="NSWO" & region=="CENTRAL") 
nswon=subset(owldata, spp=="NSWO" & region=="NORTH")
ggowc=subset(owldata, spp=="GGOW" & region=="CENTRAL")
ggown=subset(owldata, spp=="GGOW" & region=="NORTH")
ghowc=subset(owldata, spp=="GHOW" & region=="CENTRAL")
ghown=subset(owldata, spp=="GHOW" & region=="NORTH")
barrc=subset(owldata, spp=="BARR" & region=="CENTRAL")
barrn=subset(owldata, spp=="BARR" & region=="NORTH")
boowc=subset(owldata, spp=="BOOW" & region=="CENTRAL")
boown=subset(owldata, spp=="BOOW" & region=="NORTH")

#Summary Stats
#extract numbers for reporting
length(unique(boowc[boowc$yr==2024,"RouteIdentifier"])) #number of routes sampled in central Ontario
length(unique(boown[boown$yr==2024,"RouteIdentifier"])) #number of routes sampled in northern Ontario
sum(boowc[boowc$yr==2024,"Tstn"]) #number of stops sampled in central Ontario
sum(boown[boown$yr==2024,"Tstn"]) #number of stops sampled in northern Ontario

```

```{r saveindividualdata, echo = FALSE, message = FALSE}

write.csv(nswoc, "Data/2024/nswoc.csv")
write.csv(ggowc, "Data/2024/ggowc.csv")
write.csv(ghowc, "Data/2024/ghowc.csv")
write.csv(barrc, "Data/2024/barrc.csv")
write.csv(boowc, "Data/2024/boowc.csv")

write.csv(nswon, "Data/2024/nswon.csv")
write.csv(ggown, "Data/2024/ggown.csv")
write.csv(ghown, "Data/2024/ghown.csv")
write.csv(barrn, "Data/2024/barrn.csv")
write.csv(boown, "Data/2024/boown.csv")

```

#other species analysis

Here we change the code to reflect the three speices of game bird we wish to model

```{r speciesleveldata, echo = FALSE, message = FALSE}

owlss<-owls %>% dplyr::select(SamplingEventIdentifier, YearCollected, MonthCollected, DayCollected, CommonName)

rugr=subset(owlss, CommonName=="Ruffed Grouse")
rugr$nowls<-1 #for counting the number of owls 
wisn=subset(owlss, CommonName=="Wilson's Snipe")
wisn$nowls<-1
amwo=subset(owlss, CommonName=="American Woodcock")
amwo$nowls<-1


#Merge each species with the stndata dataframe to zero fill

###RUGR###
rugrroute=merge(rugr, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(rugrroute)

rugrroute$CommonName[is.na(rugrroute$CommonName)]="Ruffed Grouse" 
rugrroute <- subset(rugrroute, select = c(CommonName,YearCollected, RouteIdentifier,stn, nowls));head(rugrroute)####only the variables we want
rugrroute$nowls[is.na(rugrroute$nowls)] <- 0
summary(rugrroute)
raw_rugr<-rugrroute

rugrroute<-rugrroute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
rugrroute$spp<-"rugr"
summary(rugrroute)

###WISN###
wisnroute=merge(wisn, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(wisnroute)

wisnroute$CommonName[is.na(wisnroute$CommonName)]="Wilson's Snipe" 
wisnroute <- subset(wisnroute, select = c(CommonName,YearCollected,RouteIdentifier,stn, nowls));head(wisnroute)####only the variables we want
wisnroute$nowls[is.na(wisnroute$nowls)] <- 0
summary(wisnroute)
raw_wisn<-wisnroute

wisnroute<-wisnroute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
wisnroute$spp<-"wisn"
summary(wisnroute)

###AMWO###
amworoute=merge(amwo, stndata, by=c("SamplingEventIdentifier", "YearCollected", "MonthCollected", "DayCollected"), all=TRUE); head(amworoute)

amworoute$CommonName[is.na(amworoute$CommonName)]="American Woodcock" 
amworoute <- subset(amworoute, select = c(CommonName,YearCollected,RouteIdentifier,stn, nowls));head(amworoute)####only the variables we want
amworoute$nowls[is.na(amworoute$nowls)] <- 0
summary(amworoute)
raw_amwo<-amworoute

amworoute<-amworoute %>% group_by(CommonName, YearCollected, RouteIdentifier) %>% summarise(nowls=sum(nowls), Tstn=length(RouteIdentifier))
amworoute$spp<-"amwo"
summary(amworoute)

#combine all species again
otherroute<-rbind(rugrroute,wisnroute,amworoute)
raw_otherdata<-rbind(raw_rugr,raw_amwo,raw_wisn)
head(raw_otherdata)
otherroute$yr=as.factor(otherroute$YearCollected) #changes yr in dataframe to factor

```

```{r routeleveldata, echo = FALSE, message = FALSE}
#In the process change the datafram name to otherdata

#protocol_id 22=NORTHERN 36=CENTRAL
region <- in.data %>% dplyr::select(SamplingEventIdentifier, RouteIdentifier, protocol_id, DecimalLatitude, DecimalLongitude) %>% distinct() %>% drop_na()

#create a statation ID column
region<- region %>% separate(SamplingEventIdentifier, c("del1", "del2", "stn"), sep="-", remove=FALSE) %>% dplyr::select(-del1, -del2)

#remove leading zeros by changing route from character to integer
region$protocol_id[region$protocol_id=="22"]<-"NORTH"
region$protocol_id[region$protocol_id=="36"]<-"CENTRAL"
#change some column names
colnames(region)[which(names(region) == "protocol_id")] <- "region"
colnames(region)[which(names(region) == "DecimalLatitude")] <- "lat"
colnames(region)[which(names(region) == "DecimalLongitude")] <- "long"
#select the first stop on the route as the routes lat long
region<-region %>% filter (stn == "1") %>% dplyr::select(-stn, -SamplingEventIdentifier) %>% distinct()

#combine with the otherroute data
otherdata<-left_join(otherroute, region, by="RouteIdentifier")

#identify the regions wil missing lat long for Kathy to fix. 
#na_dropcoord<-subset(otherdata, is.na(otherdata$long))

otherdata<-otherdata %>% drop_na(lat, long)
```

#remove Covid data 2020
```{r filter years}
otherdata<-otherdata %>% filter(YearCollected!=2020)
```

Write raw datatable for MNRF Report

```{r writedata, echo = FALSE, message = FALSE}

write.csv(otherdata, "Data/2024/otherdata_2024.csv")
otherdata<-read.csv("Data/2024/otherdata_2024.csv")

```

Subset data by species and region for analysis

```{r writedata, echo = FALSE, message = FALSE}
#create a subset of rugr in CENTRAL region
rugrc=subset(otherdata, spp=="rugr" & region=="CENTRAL") 
rugrn=subset(otherdata, spp=="rugr" & region=="NORTH")
wisnc=subset(otherdata, spp=="wisn" & region=="CENTRAL")
wisnn=subset(otherdata, spp=="wisn" & region=="NORTH")
amwoc=subset(otherdata, spp=="amwo" & region=="CENTRAL")
amwon=subset(otherdata, spp=="amwo" & region=="NORTH")

```

```{r saveindividualdata, echo = FALSE, message = FALSE}

write.csv(rugrc, "Data/2024/rugrc.csv")
write.csv(wisnc, "Data/2024/wisnc.csv")
write.csv(amwoc, "Data/2024/amwoc.csv")

write.csv(rugrn, "Data/2024/rugrn.csv")
write.csv(wisnn, "Data/2024/wisnn.csv")
write.csv(amwon, "Data/2024/amwon.csv")

```
