## Setup

The setup files will get you ready for your analysis. It loads you libraries, needed tables, created directories for your outputs, and assigns some parameter values. 

Also assign max year of the analysis. 

```{r setup}

min.yr <-2018
max.yr <- 2023 #needs changes with each analysis
ID<- "dethier" #change this to your user ID. You will be prompted for a password. 

source("00-setup.R")

```

## Choose a collection to be downloaded and prepare the provincial basemap

```{r downloaddata, echo = TRUE, message = FALSE}

#list the collections
list.coll<-unique(anal.param$collection)
list.prot<-unique(anal.param$protocol_id)

#select your collection manually for data cleaning
collection <- "ONOWLS"

```

## Output Tables

Create the output tables. This only needs to be done once per provincial analysis

```{r data manip}

source("i-output.R") #Total count period output tables

```


## Data manipulations 

```{r data manip}

source("ii-datamanip.R") #Total count period

```

## Data Analysis iCAR

```{r analysis}

#Add to the trends a separate analysis for north and south
source("iii-icar.R")  #Total count period, including a covariate for number of stops per route and random effect for protocol_id to control for unexplained variation due to differences in protocol. 

```

## Data Analysis SPDE

```{r analysis}


#pull up the provincial basemap manually
qq <- rnaturalearth::ne_states(country = "canada", returnclass = "sf") %>% st_transform(epsg6703km) 
qq <- filter(qq, name %in% c("Ontario"))

source("iv-spde.R")

```

## Plot Trends 

```{r Plot trends}

##TO DO: Add to the map separate trends for North and South
source("v_plots.R")


```

## Plot Sum Stats and Visuals

```{r final outputs}

library(sf)
library(mapview)
library(ggmap) 
library(mapproj)
register_stadiamaps("1cdeb586-3f5a-4ef9-813b-02375efa9f21") 

##Plot all locations
loc <- read.csv("Output/ONOWLSEvents.csv")
loc<-loc %>% select(RouteIdentifier, latitude, longitude, survey_year) %>% distinct()

loc<-loc %>% filter(survey_year>=min.yr & survey_year<=max.yr)
test<-loc %>% select(RouteIdentifier) %>% distinct()

#View Data
sf<-st_as_sf(loc, coords = c("longitude", "latitude"), crs = 4326)
mapview(sf, map.types = "Esri.WorldImagery")


```

#View Data
LEBI_sf<-st_as_sf(in.data, 
                  coords = c("longitude", "latitude"), crs = 4326)

mapview(LEBI_sf, map.types = "Esri.WorldImagery")
